The Current State of R Tools for Nonlinear Least Squares Modeling

### 2023-07-05 RefactoringNLS/revision202307/ReviewResponses.txt

Revew Responses to first review of R Journal paper 2023-15.

We have created a text file with the content of both referees'
reports and edited our responses to comments under headings

   Response X:YY

where X indicates the referee (1 or 2) and YY is an identifying
number.

Response 0:01  We noticed a factor of 2 error in the equations
    in the discussion of the background to methods and have
    (hopefully!) corrected them.

@@

What was called QPE (quantitative programming environment), then “The New 
S Language”, then “S3” represented a huge step forward for nonlinear 
modeling because it could represent model functions as expressions 
(i.e. abstract syntax trees) that could be manipulated in the system 
itself. This allowed for symbolic differentiation.

The original implementation of the deriv function for symbolic 
differentiation was written in S. When porting deriv to R in the 
mid-1990’s Ross Ihaka moved quite a bit of the functionality to C code 
because it was comparatively easy to do so and, on the hardware of the 
day, the C implementation provided a considerable speed advantage. The 
speed of the differentiation itself isn’t usually a problem - it’s the
common-sub-expression elimination that can be the bottleneck.

This legacy is also the reason that the Jacobian matrix is the "gradient" 
attribute of the returned value of the model function. For a nonlinear 
least squares problem, where the model function returns a vector, this 
deriv attribute should be called "Jacobian" but the version for the 
scalar objective, for which the vector of partial derivatives is the 
gradient, was written first. Because deriv only has access to the 
expression and not to the model frame it is not possible to determine 
if the value will be a scalar or a vector and the name of the attribute 
was left as "gradient". A further complication is that numerical values 
in S and R are always stored as vectors, with the understanding that a 
scalar is represented as a vector of length 1. Even if it were desirable 
to distinguish between scalars and vectors, internally it is impossible 
to determine if a vector of length 1 is intended to be a scalar or to 
be a vector that just happens to have length 1.

Response 1:07 Thanks for explaining this so clearly. We have added a
    note in section "Estimation of models specified as formulas".

@@

Some of the discussion in the “Jacobian code in selfStart models” section 
(p. 9) of this paper claims that the plinear fit followed by a fit using 
the default approach would be difficult to maintain and could be very slow
but that has not been the case. It is known (and, in fact, stated in the 
documentation for SSlogis) that the call to the default nls algorithm will 
always declare convergence immediately.

Response 1:03 Can we have a pointer to the documentation of the immediate 
      convergence? We have checked the documentation ?SSlogis, as well as 
      selfStart and getInitial, but do not see a mention in R 4.3.1.

      Interim response: We have reworded the paragraph starting
	"In considering `selfStart` models,"
      to reflect the comments of the referee and to provide a reference
      to the source of the starting approximations we used in SSlogisJN().

@@

Response 2:03 "port" cases added. The "appears unfinished" comment in the 
       documentation led us to avoid its use. This is the sort of issue we 
       would very much like to see resolved, since "port" is the only tool
       for bounds constraints.

Also, the model setup (Logistic3T) for which gsl_nls() fails can easily be 
made to converge correctly by changing either the algorithm or the scaling 
method, see the Illustrating code section. My advice is to provide a more 
comprehensive comparison of the available NLS solvers, or to pick a different
data example that better highlights the benefits of nlsr. For instance, trying 
to solve the BoxBOD regression problem 
(https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), 
nls() fails with starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton 
and Port algorithm, whereas nlsr::nlxb() converges correctly.

Response 2:04 We have added a paragraph

	We consider the code inside package `nlsr` to be competently coded, 
	with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
	in common with `minpack.lm` and `glsnls`. The success rate is high,
        but `nlsr` has its premature terminations failures too. Our principal 
        message is that we obtain diagnostic information in the form of the 
        singular values of the Jacobian that allow us to try alternative starts, 
        algorithms or programs in cases where there are indications of early termination.

   We appreciate the pointers on other examples. However, a substantive comparison
   would make for a very long and likely boring paper. The BoxBOD example is, in
   many ways, similar to the Hobbs example in its nasty characteristics. 
   Experiences with these problems have led to an opinion (i.e., open to debate)
   as to the usefulness of such extreme cases in measuring performance. We want 
   to underline a point of view that good diagnostics are more helpful than a 
   "usually succeeds" level of reliability.

@@

The section Returned results of nls() and other tools discusses the complexity 
of S3-objects of class nls, illustrated by the retrieval of the input data 
from an nls-object. It seems ill-advised to run eval(parse(text=result$data)), 
since it evaluates the name of the data object in the user’s environment, 
which may not exist. Instead, one should use model = TRUE to store the input 
data in the returned nls-object. This behavior is consistent with lm(), so 
perhaps less confusing than currently presented in the manuscript. The 
authors mention that nlxb() returns a much simpler structure of 11
items in one level, but I did not manage to find the input data object in 
the nlsr-object returned by nlxb().

Response 2:05 The section has been reworded to reflect the comments. I (JN) 
   confess to having used `lm()` only a few times in a whole career, so
   missed the parallel to that function. My work has focused on developing 
   computational tools or diagnosing extreme cases where conventional 
   statistical calculations seem to have misfired, and also do not work 
   exclusively in R.

The section Functional specification of problems mentions the use of 
nlsr::nlfb() to solve nonlinear least squares problems using a function 
to define the residuals. Since nlsr::nlfb() only has a notion of the 
residuals, how is it able to disentangle model and response, for the 
purpose of e.g. generating model predictions? The authors mention that 
the function wrapnlsr() can be used to return an object of class nls, 
but this does not seem to work for problems defined through a residual 
function as in nlsr::nlfb()? If indeed the case, how does one evaluate 
e.g. model predictions, parameter confidence intervals, etc. when using 
nlsr::nlfb()? This remains unclear to me.

Response 2:06 The referee's comments point out that we have not explained
    clearly enough the important differences that arise between modeling
    and minimizing a nonlinear sum of squares. We have added a short 
    exposition using the Rosenbrock problem. The difference in context
    is important and we thank the referee, as other readers would surely
    have had similar concerns. We have also added a comment to the 
    "Problem setup" section to note that `nls()` is not suited to this
    approach. Also a minor amendment to indicate `wrapnlsr()` is intended
    for problems specified by formula.

@@

    We have added two small comments on recommended usage of bounds. One
    is to control for wild excursions in the optimization trajectory, as 
    we believe this is sensible when bounds constraints are available, 
    as in nlsr::nlxb. We also note that in cases where a simple and 
    obvious transformation is available, such as squaring a parameter to 
    ensure positivity, this is likely a better approach than imposing bounds.
    However, when there are more general bounds, transformations become more
    tricky to apply, and may de-scale a problem. We have considered the
    transfinite function used in CRAN package `adagio`, but consider that
    this needs more extensive investigation before recommending in general.

5. The section Programming language makes the argument that keeping to a 
single programming language can allow for easier maintenance and upgrades, 
and that the performance penalty for using code entirely in R has been 
much reduced in recent years. I agree that there is not much of a 
performance penalty when solving small problems, such as the data examples 
included in the article. However, when solving problems using a large amount 
of data or a large number of parameters, I have experienced that the run-time 
(as well as memory usage) can become an actual bottleneck. As an example, 
consider the Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) 
included in the Illustrating code section.

Response 2:08  In that we are in general agreement with this comment, we have added
    some clarification to advise that it is worthwhile being aware of how long
    calculations are taking.

    The referee's commends ant Illustrating Code led us to try the Pen1 problem
    (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
    to which one of us recently added Hessian code). This showed that the R code
    of nlsr::nlfb() was generally slower by a small factor than the compiled 
    Fortran of minpack.lm::nls.lm. However, the all-R code optimx::Rcgmin was 
    much faster for larger problems -- it does not build a matrix!

	## On HP21 (Ryzen 7, 16GB)

	> print(TT)  # median times
	       n       nlfb    nls.lm     Rcgmin
	[1,]  50   15.02918   7.95267   3.399167
	[2,] 100   96.74012  40.05409  96.431357
	[3,] 150  319.44073 127.58728   6.143749
	[4,] 200  605.29889 290.00972   6.109964
	[5,] 250 1373.51954 532.36149 182.678074
	[6,] 300 2732.06920 874.97992  12.007694

	> print(MM)  # minimum times
	       n       nlfb     nls.lm     Rcgmin
	[1,]  50   14.69422   7.895639   3.345110
	[2,] 100   92.10252  39.718163  94.155118
	[3,] 150  281.62738 125.972404   6.042982
	[4,] 200  595.52840 284.833648   5.816762
	[5,] 250 1300.00651 526.563457 182.678074
	[6,] 300 2662.15242 869.606156  11.600069

### 2023-07-05 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
updates to rev1

?? check wording

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

It is also possible to provide R functions for the residual and Jacobian. 
This is usually more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.

@@

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see `nls()` does poorly on the example problem. To save page space,
we use 1-line result display functions from package `nlsr`, namely `pnls()` 
and `pshort()`.

@@

Here we see the infamous "singular gradient" termination message of `nls()`.

However, `nls()` also offers the option of using the "port" library algorithm
choice (@FoxHallSchryer1978). There is a warning in the `nls()` documentation that
this "appears unfinished", but it is successful in solving all three cases.

@@

@@

The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a very 
large sum of squares, which may suggest that these programs have failed. 
Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
offer singular values, we need to extract the Jacobian and compute its
singular values. The following script shows how to do this, using as Jacobian what is called
the `gradient` element in the returned solution for these solvers.

@@

We consider the code inside package `nlsr` to be competently coded, with
the advantage over `nls()` of the Levenberg-Marquardt
stabilization in common with `minpack.lm` and `glsnls`. However, our principal 
message is that we obtain diagnostic information in the form of the singular values
of the Jacobian that allow us to try alternative starts, algorithms or programs in 
cases where there are indications of early termination.

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case, is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note function `nlsr::pnlslm()` for a 1-line display of
the results of `minpack.lm::nls.lm()`.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnlslm(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnls(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnls(hobgsl) # using analytic Jacobian
```

@@

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()`, and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. The object of class `nls` allows for 
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias. 
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

@@

with suggested addition of small perturbations to the data. This admits 
`nls()` could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Fortunately, a small adjustment
to the "termination test" for the **program**, rather than for the
"convergence" of the underlying **algorithm**, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a 
small number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
since it was introduced. The default value 
gives the legacy behaviour. This improvement has been in the R distributed code 
since version 4.1.0.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

The complexity of this object is a challenge to users. Let us use the
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. 
If the user's environment exists, which it may not, we could see this data 
via the syntax:
```
eval(parse(text=result$data))
```
However, a better approach is to make the call to `nls()` with `model=TRUE`, 
which returns an element 
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Tools that produce a class `nls` output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
`nlsr` computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian 
are actually computed in the `print` and `summary` methods for the result. 
These two approaches
lead to different consequences for performance and how features are
provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in `nlsr` is likely of value for workers to copy and modify
for customized tools.

@@

Using the "gradient" attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the **residual** function as well,
so that the call to `nlsr::nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

Analytic Jacobian code can be provided to all the solvers discussed.
Most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but their construction is non-trivial. A number 
of such models are included with base R in `./src/library/stats/R/zzModels.R`,
with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
There are also some in the now-archived package `NRAIA`.
These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
that defines each model, and these Jacobians are often the analytic forms.

The `nls()` function, after
computing the "right-hand side" or `rhs` of the residual, checks to see if the
"gradient" attribute is defined, otherwise using `numericDeriv()` to compute a 
Jacobian into that attribute. This code is within the `nlsModel()` or
`nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a (full) solution. It then passes the solution coefficients to the default 
algorithm unnecessarily. Moreover, the implicit double call is, in our view, prone to creating 
errors in code maintenance. 
To provide simpler starting parameters, the function `SSlogisJN`
is now part of the package `nlsr`, but is most useful for `nls()`.

@@

We note that the analytic expressions for the Jacobian ("gradient") in the `SSLogis` 
function and others save quantities by default in "hidden" variables, i.e., with 
names preceded by ".".  A referee has indicated this behaviour is to avoid name 
conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
command, so users need to remember to use `ls(all.names=TRUE)` to see which quantities
are defined, for example, in developing their own selfStart model via copy and edit.

@@

One of the referees pointed out that nlsr::nlfb() is quite slow on a
"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
However, we would suggest that problems in large numbers of parameters
will be problematic for Gauss-Newton methods because they create large
computational matrices for the Jacobian or its inner product. The method
"Rcgmin" from package `optimx` (@p-optimx) can be orders of magnitude
faster, but is entirely written in R. Note that "Rcgmin" is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. It should be noted
that timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods.

>>>>

@@

Let us consider there are $n$ parameters and $m$ residuals. 
The loss function is 
$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
The gradient of $S(p)$ is
$$ g = 2 * J' r$$
where the Jacobian $J$ is given by elements
$$ J_{i,j} = \partial r_i / \partial p_j $$
and the Hessian is defined by elements
$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
If we expand the Hessian for nonlinear least squares problems, we find
$$  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
Let us use $D_{i,j}$ for the elements of the second term of this expression.
What is generally called **Newton's method** for function minimization tries
to set the gradient to zero (to find a stationary point of the function $S(p)$).
This leads to **Newton's equation**
$$ H \delta  = -g $$
Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
to $p + \delta$ and iterate, hopefully to converge on a solution. 
Applying this to a sum of squares problem gives
$$ H \delta / 2 = (J'J + D) \delta = - J' r$$
In this expression, only the elements of $D$ have second partial derivatives. 
Gauss, attempting to model planetary orbits, had small residuals, and noted
that these multiplied the second partial derivatives of $r$, so he approximated
$$ H / 2 \approx J' J$$
by assuming $D \approx 0$. This results in the Gauss-Newton method where we
solve 
$$ J' J \delta = - J' r$$
though we can avoid some loss of accuracy by not forming the inner product matrix
$J' J$ and solving the linear least squares matrix problem 
$$ J \delta \approx -r $$
by one of several matrix decomposition methods.

@@

It is also possible to provide R functions for the residual and Jacobian.
`nls()` is not well-suited to using such a formulation, but `nlsr::nlfb()`
and `minpack.lm::nls.lm()` work specifically with such a formulation.
This is usually more work for the user if the formula setup is possible.
To illustrate, we show the functions for the unscaled 3 parameter logistic.

@@

Using a formula specification was a principal advantage made
with `nls()` when it became available in S sometime in the 1980s. It uses a
Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
search. This works very efficiently as long as $J$ is not ill-conditioned.
Below we see the default `nls()` algorithm does poorly on the example problem. 
To save page space, we use 1-line result display functions from package `nlsr`, 
namely `pnls()` and `pshort()`.

@@

Here we see the infamous "singular gradient" termination message of `nls()`. 
Technically it is the Jacobian that is singular. A
referee noted that R in this case cannot distinguish scalar and vector functions,
which would yield partial derivatives that are a gradient or Jacobian respectively.

`nls()` also offers the option of using the "port" library algorithm
choice (@FoxHallSchryer1978). There is a warning in the `nls()` documentation that
this "appears unfinished", but it is successful in solving all three cases.

@@

While default `gsl_nls()` gives much the same results as `nlsLM()`, uncommenting the
calls to the alternative algorithms shows it can get the global minimum with any of
these choices.

@@

The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a 
very large sum of squares, which may suggest that these programs have failed. 
Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
offer singular values, we need to extract the Jacobian and compute its
singular values. The following script shows how to do this, using as Jacobian 
the `gradient` element in the returned solution for these solvers.

@@

We consider the code inside package `nlsr` to be competently coded, 
with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
in common with `minpack.lm` and `glsnls`. The success rate is high,
but `nlsr` has its premature terminations too. Our principal message is that we obtain 
diagnostic information in the form of the singular values of the Jacobian 
that allow us to try alternative starts, algorithms or programs in cases 
where there are indications of early termination.

@@

We illustrate how to solve nonlinear least squares problems using
a function to define the residual. Note that `gsl_nls()` requires
a vector `y` that is the expected value of what we have called the
residual, but in this case, is actually the model. `gsl_nls` uses 
a numerical approximation for the Jacobian if the argument `jac` 
is missing. Note function `nlsr::pnls()` for a 1-line display of
the results.

```{r exhobbsfn, eval=TRUE}
hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
pshort(hobnlfb) # use print(hobnlfb) for more detail
hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
pnls(hobnlm)  
hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
pnls(hobgsln)
hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
pnls(hobgsl) # using analytic Jacobian
```

@@

The output object of `nlxb()` is smaller than the class `nls` object returned
by `nls()`, `nlsLM()`, and `gsl_nls()`. Package `nlsr` emphasizes the solution 
of the nonlinear least squares problem rather than the estimation of a nonlinear 
model that fits or explains the data. The object of class `nls` allows for 
a number of specialized modeling and diagnostic extensions. For compatibility, 
the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias,
which is intended for problems specified as formulas.
This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
`nls` object. Unless particular modeling features are needed, the use of 
`wrapnlsr()` is unnecessary and wasteful of resources.

@@

with suggested addition of small perturbations to the data. This admits 
`nls()` could not solve well-posed problems unless data is polluted
with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be 
approached with nonlinear least squares. Fortunately, a small adjustment
to the "termination test" for the **program**, rather than for the
"convergence" of the underlying **algorithm**, fixes the defect. The test
is the Relative Offset Convergence
Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
function by its current value. If the loss function is very small, we are close
to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
additional parameter `scaleOffset` with a default value of zero. Setting it to a 
small number -- 1.0 is a reasonable choice -- allows small-residual problems 
(i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
since it was introduced. The default value 
gives `nls()` its legacy behaviour. This improvement has been in the R distributed code 
since version 4.1.0.

Additional termination tests can be used. `nlsr` has a **small sum of 
squares** test (`smallsstest`) that compares the latest evaluated sum of squared
(weighted) residuals to `e4` times the initial sum of squares, where
`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.

@@

The complexity of this object is a challenge to users. Let us use the
`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. 
If the user's environment exists, which it may not, we could see this data 
via the syntax:
```
eval(parse(text=result$data))
```
However, since the evaluating environment might not exist (thanks to a referee
for noting this possibility) a better approach is to make the call to `nls()` 
with `model=TRUE`, which returns an element 
`model` which contains the data, and we can list its contents using:
```
ls(result$model)
```
and if there is an element called `xdata`, then it can be accessed as
`result$model$xdata`.

@@

Tools that produce a class `nls` output object create a rich set of functions
and structures that are then used in a variety of modeling tasks, including the
least squares solution. By contrast,
`nlsr` computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the Jacobian 
are actually computed in the `print` and `summary` methods for the result. 
These two approaches
lead to different consequences for performance and how features are
provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
data and programs was at a ridiculous premium in the small computers of the era.
Thus the code in `nlsr` is likely of value for workers to copy and modify
for customized tools. Coding in R is intended to support such uses.

@@

Using the `gradient` attribute of the output of the Jacobian function to hold
the Jacobian matrix lets us embed this in the **residual** function as well,
so that the call to `nlsr::nlfb()` can be made with the same name used for 
both residual and Jacobian function arguments. This programming trick saves a 
lot of trouble for the package developer, but it can be a nuisance for users 
trying to understand the code.

@@

Analytic Jacobian code can be provided to all the solvers discussed.
Most `selfStart` models that automatically provide starting parameters
also include such code. There is documentation in R
of `selfStart` models, but their construction is non-trivial. A number 
of such models are included with base R in `./src/library/stats/R/zzModels.R`,
with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
There are also some in the now-archived package `NRAIA`.
These provide the Jacobian in the `gradient` attribute of the "one-sided" formula
that defines each model, and these Jacobians are often the analytic forms.

The `nls()` function, after
computing the "right-hand side" or `rhs` of the residual, checks to see if the
`gradient` attribute is defined, otherwise using `numericDeriv()` to compute a 
Jacobian into that attribute. This code is within the `nlsModel()` or
`nlsModel.plinear()` functions. The use of analytic Jacobians
almost certainly contributes to the good performance of `nls()` on `selfStart` 
models.

@@

In considering `selfStart` models, we noted that the base-R function `SSlogis`
is intended to solve problem **Logistic3T** above.  When this function is used
via `getInitial()` to  find
starting values, it actually calls `nls()` with the 'plinear' algorithm 
and finds a (full) solution. It then passes the solution coefficients to the default 
algorithm which converges in one iteration and builds the standard output. 
We found the implicit double call a challenge to unravel. 
A more conventional approach to starting parameters, following the lead of @Ratkowsky1983, 
is in the function `SSlogisJN`, now part of the package `nlsr`.

@@

We note that the analytic expressions for the Jacobian (`gradient`) in the `SSLogis` 
function and others save quantities by default in "hidden" variables, i.e., with 
names preceded by ".".  A referee has indicated this behaviour is to avoid name 
conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
command, so users need to remember to use `ls(all.names=TRUE)` to see which quantities
are defined, for example, in developing their own selfStart model via copy and edit.

@@

One of the referees pointed out that nlsr::nlfb() is quite slow on a
"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
However, we note that problems in large numbers of parameters
will be problematic for Gauss-Newton methods because they create large
computational matrices for the Jacobian or its inner product. The method
"Rcgmin" from package `optimx` (@p-optimx) can be much 
faster, despite being entirely written in R, and is a much more
reliable conjugate gradient minimizer than method "CG" of `optim()`. One
of us created both, but now deprecates the latter. 
Timings of CG methods are more sensitive to small differences in
setup and arithmetic than Gauss-Newton methods.
